Select-based random access to variable-byte encodings

### 2020-03-19 gradu/msc_english_contents.tex
new stuff

The shown $rank$ method by Brisaboa et al. can also effectively decompress the next value but some modifications are needed. When fetching a longer subarray, the goal becomes to minimize $rank$ calls. 
If an additional pointer array is stored during subarray decompression, only one $rank$ call for each accessed block level is needed. This can be observed from Figure~\ref{bris_ds}; the blocks of subarray
are located next to each other in different block levels.

@@

\hline
%
\end{tabular}
\end{table}

The results show that the proposed $select$ method outperforms $rank$ when the data set contains several integers that are encoded into more than one block. The difference is explained 
with how $rank$ and $select$ use the data structure. $Select$ is always called once, while $rank$ is called once for each block beyond the first. The 8-bit versions tend to perform better than the 
4-bit versions. This is because 8-bit version stores more data in one block and less block readings are required. However, 4bit versions achieve better compression in data sets containing small integers.
Select versions do not depend as much on the block size, because all the blocks in 
the compressed element are read at one go.

@@

Interestingly the 4-bit version of $select$ performs better than the 8-bit version with some data sets, even though 4-bit version should execute at least the same commands as the 8-bit version 
in every scenario. This is most likely due to cache, because in the 4-bit version the stored data set is smaller.

@@

 - see how data length percentage matters (do datasets with k/8 long numbers).

@@

 - See Ahn \& Moffat 2005; Lemire et al. 2018

 - subarray access(?)

@@

 - get compression sizes of 4 and 8 bit versions

>>>>

The shown $rank$ method by Brisaboa et al. can also effectively decompress the next value but some modifications are needed. When fetching a longer subarray, the goal becomes to minimize $rank$ calls. 
If an additional pointer array is stored during subarray decompression, only one $rank$ call for each accessed block level is needed. This can be observed from Figure~\ref{bris_ds}; the blocks of a subarray
are located next to each other in different block levels.

@@

\hline
\end{tabular}
\end{table}

The results show that the proposed $select$ method outperforms $rank$ when the data set contains several integers that are encoded into more than one block. The difference is explained 
with how $rank$ and $select$ use the data structure. $Select$ is always called once, while $rank$ is called once for each block beyond the first. Figure~\ref{graph:data_comparison} shows average times
of one million $rank$ and $select$ calls with different average block lengths. The effect of multiple $rank$ calls is visible, as is the single $select$ call behavior.

The 8-bit versions tend to perform better than the 4-bit versions. This is because 8-bit version stores more data in one block and less block readings are required. However, 4bit versions achieve 
better compression in data sets containing small integers. Select versions do not depend as much on the block size, because all the blocks in the compressed element are read at the same time. 
Interestingly the 4-bit version of $select$ performs better than the 8-bit version with some data sets, even though 4-bit version should execute at least the same commands as the 8-bit version 
in every scenario. This is most likely due to cache, because the stored data set is smaller in the 4-bit version.

@@

\begin{figure}[ht]
\centering
\begin{tikzpicture}
\begin{axis}[
    xlabel={average integer byte length},
    ylabel={ms per million queries},
    xmin=1, xmax=8,
    ymin=0, ymax=800,
    xtick={1,2,3,4,5,6,7,8},
    ytick={0,100,200,300,400,500,600,700,800},
    legend pos=north west,
    ymajorgrids=true,
    grid style=dashed,
]

\addplot[
color=blue,
mark=square,
]
coordinates {
(1,31.90)(2,111.00)(3,205.40)(4,304.20)(5,405.80)(6,513.90)(7,616.90)(8,725.10)
};
\addplot[
color=green,
mark=triangle,
]
coordinates {
(1,32.90)(2,145.30)(3,271.20)(4,399.00)(5,528.20)(6,666.80)(7,767.40)(8,902.00)
};
\addplot[
color=red,
mark=ball,
]
coordinates {
(1,202.40)(2,219.70)(3,228.70)(4,251.60)(5,261.80)(6,269.60)(7,275.70)(8,284.30)
};
\addplot[
color=black,
mark=line,
]
coordinates {
(1,218.60)(2,255.20)(3,273.00)(4,282.90)(5,296.00)(6,306.10)(7,317.30)(8,327.50)
};

\legend{bris8, bris8v5, select8, select4 }

\end{axis}
\end{tikzpicture}
\caption{Rank and select performances} \label{graph:data_comparison}
\end{figure}

Figures \ref{table:results8bit} and \ref{table:results4bit} show memory size requirements for $rank$ and $select$ data structures. The data sets consist of 50 million 64-bit numbers and 
the compressed size is a sum of continuation bit array and the data blocks.

The static memory requirement of $select$ is easily noticeable, 
as is the relation between the memory requirement of $rank$ and the total number of blocks. Data structures in versions with smaller block size naturally take more space, but smaller block 
size allows a better data compression in some data sets and therefore a smaller total size. This can be seen when comparing the two forementioned figures: All 4-bit versions of the algorithms require
more space than the 8-bit versions, but the compressed size of the data is significantly smaller in the bottom two data sets. The data structure memory usage for the bottom row is similar in both versions 
due to the fact that the continuation bit array is exactly the same.

\begin{table}
\centering
\begin{tabular}{l|| c c c c } 
dataset   & compressed size & bris8  & bris8v5 & select8  \\
 \hline \hline 
all       & 168.4MB         & 4.67MB & 1.17MB   & 1.54MB    \\
twolarge  & 112.4MB         & 3.12MB & 781kB    & 1.48MB    \\
onelarge  &  63.2MB         & 1.76MB & 439kB    & 1.43MB    \\
onlysmall &  56.3MB         & 1.56MB & 391kB    & 1.43MB    \\

\hline
%
\end{tabular}
\caption{Memory requirement for 50M 64-bit numbers 8bit blocks, smaller is better.\label{table:results8bit}}
\end{table}

\begin{table}
\centering
\begin{tabular}{l|| c c c c } 
dataset   & compressed size & bris4  & bris4v5 & select4  \\
 \hline \hline 
all       & 177.7MB         & 7.65MB & 1.91MB  & 1.63MB \\
twolarge  & 116.1MB         & 4.66MB & 1.14MB  & 1.54MB \\
onelarge  & 42.5MB          & 2.12MB & 531kB   & 1.44MB \\
onlysmall & 31.3MB          & 1.56MB & 391kB   & 1.43MB \\

\hline
%
\end{tabular}
\caption{Memory requirement for 50M 64-bit numbers 4bit blocks, smaller is better.\label{table:results4bit}}
\end{table}

@@

@@

 - subarray access

@@
